---
title: "Quality control of Spectronaut proteomic data (Both neat and depleted)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: 
      collapse : false
    toc_depth: 3
    code_folding: hide
    number_sections: true
    css: report_aux/styles.css
# bibliography: citations_in_report.bib
knit: (function(input, ...) rmarkdown::render(input, ..., output_dir= "../reports"))
---

```{r Setup, echo=FALSE, include=FALSE}
source("report_aux/rmd_usual_setup.R")
# Pre-loaded packages for report only
library(kableExtra)
library(janitor)
```

```{r Header, purl=TRUE}
stopifnot(getRversion() >= "4.1.0")   # R base pipe

## Pre-loaded packages
# Please note that there are possibly more packages employed by direct calling
# with double-colon '::' or triple-colon ':::'.
library(tidyverse)

# File neames
fn <- list(
  i = list(   # input
    nt01 = "../data/s1_neat_Spectronaut.v01.RData",
    dp01 = "../data/s1_depl_Spectronaut.v01.RData"
  ),
  pr = list(  # for presentation here 
    run_neat = "../data/raw_internal/2021-11-19/NeatPlasma_MS_run_conditions.xlsx",
    injt = "../data/s1_neat_injection.v01.RData",
    c01 = "../data/s1-clinic.v01.RData"
  ),
  o = list(   # output
    interim = "../data/interim/Spectronaut_after_trimming.RData"
  ),
  check = "../data/s1_Spectronaut.v02.RData"
)
# Brief check if all files exist
stopifnot(all(file.exists(unlist(fn$i), dirname(unlist(fn$o)))))
```

## Two data sets

The proteomic data of neat plasma and depleted plasma from a set of individuals (of two cohorts) were obtained. 
The data of the two types were attached by matching sample IDs, setting apart the same protein of the two types.

```{r Load data, purl=TRUE}
# Load and combine proteomic data
e <- lapply(
  setNames(c("nt01", "dp01"), c("Neat", "Depleted")), 
  \(.x) {
    load(fn$i[[.x]], e <- new.env())
    e
  }
)
proteins <- imap_dfr(e, ~mutate(.x$proteins, prep_type = .y)) |> 
  mutate(protein_type_id = paste(id, prep_type, sep = "-")) 
spct <- imap_dfr(e, ~mutate(.x$spct, prep_type = .y)) |> 
  mutate(protein_type_id = paste(protein_id, prep_type, sep = "-")) |> 
  separate(sampleid, c("subjid", "visit_nr"), sep = "-", remove = FALSE)
prep_types <- set_names(unique(spct$prep_type))
rm(e)
```

```{r FUN - long_to_matrix & to_matrix_by_prep_type, purl=TRUE}
#' Transform a long data frame to a matrix. Row names were filled with given
#' values in `row`.
#'
#' @param x a longitudinal data frame that has those three columns given by
#' `row`, `col` and `value`
#' @param row,col which column will be the rows/columns of the output
#' @param value the column of the values of the output matrix
#'
#' @return a matrix
long_to_matrix <- function(x, 
                           row = sampleid, 
                           col = protein_type_id, 
                           value = LogIntensities) {
  pivot_wider(
    data = x,
    id_cols = {{row}}, 
    names_from = {{col}},
    values_from = {{value}}
  ) |> 
    column_to_rownames(as.character(enexpr(row))) |> 
    as.matrix()
}

#' Transform to a list of matrices, which were divided by `prep_type`
#'
#' @param x a data frame
#' @return a list of matrices
to_matrix_by_prep_type <- function(x) {
  split(x, x$prep_type) |>     # by prep_type
    lapply(long_to_matrix)
}
```


### Dimension

* Number of samples : `r n_distinct(spct$sampleid)`
* Number of proteins : `r n_distinct(spct$protein_type_id)`

#### Samples

```{r}
sids <- map(prep_types, ~unique(spct$sampleid[spct$prep_type == .x]))
```

The samples only analyzed after immuno-depletion were 
`r knitr::combine_words(setdiff(sids$Depleted, sids$Neat), before = "\u0060")`,
while those without the depletion only were
`r knitr::combine_words(setdiff(sids$Neat, sids$Depleted), before = "\u0060")`.

```{r, results='markup'}
# number of samples / subject
per_subj <- distinct(spct, sampleid, subjid, visit_nr) |> 
  add_count(subjid) |> 
  rename(n_per_subj = n)    # tabyl also create 'n'

per_subj_id <- per_subj |> 
  group_by(n_per_subj) |> 
  summarise(subjs = paste(subjid, collapse = ", ") |> str_trunc(50), .groups = "drop")
  
per_subj |>
  tabyl(n_per_subj, visit_nr) |>
  left_join(per_subj_id, by = "n_per_subj") |> 
  adorn_title()
```

As one subject `194532` was analyzed for one time-point and by one preparation method (with depletion) only, all the data of the sample was removed. 

```{r Remove the orphan sample, purl=TRUE}
spct <- spct |> filter(sampleid != "194532-T1")
```

#### Proteins

```{r Venndiagram-proteins by type}
measured_by <- proteins |> 
  mutate(value = TRUE) |> 
  pivot_wider(id_cols = id, names_from = prep_type, values_fill = FALSE)

ggplot(measured_by) +
  ggvenn::geom_venn(aes(A = Neat, B = Depleted)) +
  coord_fixed() +     # avoid oval shapes
  theme_void()       # no frame
```


## Quality control

_This chapter describes how the provided Spectronaut proteomic data was examined and processed for quality control._
_Tables and figures are presented reflecting intermediate status during the preprocessing of the data._

```{r FUN - prop_scatter_hist & attach_clinc}
#' Scatter plat & histogram of proportion
#'
#' @param df_prop a data frame with `prop` and `wise`. Usually an output of
#'   `cond_prop`.
#' @param x the value to show in x-axis. The `prop` will be presented in y-axis.
#' @param color the color aesthetic for those plots
#' @param title title
#' @importFrom expss var_lab
#' @importFrom rlang as_label
#' @importFrom grid textGrob gpar grid.arrange
prop_scatter_hist <- function(df_prop, x, color, title = NULL) {
  ylab <- expss::var_lab(df_prop$prop)  # variable label for y-axis label
  
  # scatter plot
  scatter_ms <- ggplot(df_prop) +
    geom_point(aes(x= {{x}}, y= prop, color = {{color}})) +
    facet_grid(wise ~ .) +
    theme(strip.text.y = element_blank()) +   # remove the labels of the facets
    ylab(if (is.null(ylab)) "Proportion" else ylab) +
    xlab(paste(rlang::as_label(enquo(x)), "of `LogIntensities`"))

  # histogram       
  hist_ms <- ggplot(df_prop, aes(x= prop)) +
    geom_histogram(aes(y = ..density.., fill = {{color}}), bins = 20, 
                   position = "identity", alpha = 0.3) +
    geom_density(aes(color = {{color}})) +
    coord_flip() +
    facet_grid(wise ~ .) +
    labs(x= "", y= "Density") +
    guides(fill = "none", color = "none")

  # both plots in one frame
  top_title <- grid::textGrob(title, gp = grid::gpar(fontsize = 20))
  gridExtra::grid.arrange(
    scatter_ms, hist_ms,
    ncol= 2, nrow= 1, widths= c(4, 2),
    top = if(is.null(title)) NULL else top_title
  )
}
```


### Summary

* No redundant nor exactly repeated value
* Poor quality data was removed.
* Imputation : replaced with a randomly chosen value between one and the minimum
* Normalization : Cyclic Loess, neat and depleted samples separately
* Inverse correlation between missingness and signal intensity was observed in proteins but not in samples. 
It reflects some values of our data were missing not at random (MNAR) by lower-limit of detection. 
* No indication for patient (or proteoform) dependent missingness was found. 

### Redundancy

No pairs of protein profiles were perfectly correlated.
This check-up was added to detect if any group of values labelled with multiple proteins (or features) redundantly. 

```{r Redundancy}
# correlation between proteins
corr <- to_matrix_by_prep_type(spct) |>
  lapply(\(ea) {
    out <- cor(ea, use = "pairwise.complete.obs")   # handle missing values
    
    # the number of values except NA in the columns of a pair
    na_or_not <- !is.na(ea)
    only_two <- (t(na_or_not) %*% na_or_not) < 3
    out[only_two] <- 0    # meaningless two vs. two correlation
    out
  })

# confirm no perfectly correlated proteins
stopifnot(
  all(corr$Depleted[upper.tri(corr$Depleted)] < 1),
  all(corr$Neat[upper.tri(corr$Neat)] < 1)
)
```

### Exactly repeated values

```{r}
stopifnot(anyDuplicated(spct$LogIntensities) == 0)
```

There was not any exactly repeated values.
This check-up is for identification of such unexpected values in quantitative proteomic data. 

### Distribution of signals

Global distribution of protein signals is visualized. 

```{r}
ggplot(spct, aes(LogIntensities)) +
  geom_histogram(aes(y = after_stat(density), fill = prep_type), 
                 bins = 40, position = "identity", alpha = 0.5)
```

```{r, fig.height=4}
ggplot(spct) +
  geom_qq(aes(sample = LogIntensities)) +
  facet_wrap(~prep_type)
```

The `LogIntensities` were as below.

```{r, results='markup'}
summary(spct$LogIntensities)
```

#### Zoom-in around 0

The distribution of the signals around 0 (below 5) is presented below. 
Some not many values close to zero were identified. 

```{r Plot - LogIntensities < 5, fig.height=4}
lb <- floor(min(spct$LogIntensities) * 5) / 5
ggplot(filter(spct, LogIntensities < 5)) +
  geom_histogram(aes(LogIntensities, fill = prep_type), 
                 breaks = seq(lb, 5, by = 1/5),
                 position = "identity", alpha = 0.5) +
  ggtitle("LogIntensities < 5")
```

#### Log intensity ≤ 0

```{r}
tmp <- sum(spct$LogIntensities <= 0)
```

There were `r tmp` (`r round(tmp/dim(spct)[1L] * 100, 5)`%) values below zeros in the given data. 
All of them were from depleted samples.
Because the following imputation chooses 0 as the lower bound of imputed values, those values below zero are replaced with zero. 

```{r * Replace below zero *, purl=TRUE}
spct <- spct |> 
  mutate(LogIntensities = if_else(LogIntensities < 0, 0, LogIntensities))
```

```{r Plot - LogIntensities < 5 after replacement, fig.height=4}
ggplot(filter(spct, LogIntensities < 5)) +
  geom_histogram(aes(LogIntensities, fill = prep_type), 
                 breaks = seq(lb, 5, by = 1/5),
                 position = "identity", alpha = 0.5) +
  ggtitle("LogIntensities < 5 (After replacement)")
```

### Missing measures

```{r FUN - cond_prop & cond_prop_by_prep_type, purl=TRUE}
#' Compute the proportion of a condition, together with other summary statistics
#'
#' @param mat a input matrix
#' @param cond_fun a function that returns a logical vector per row/column. The
#'   proportion of TRUE is stored in the `prop` of the output.
#' @param aggre_funs a list of functions with names, each of which compute a
#'   summary statistic per row/column, e.g. median. Every function should accept
#'   the argument `na.rm`. As the default, `na.rm` is set to TRUE.
#' @param wises a vector of two characters, which tell what rows/columns
#'   contain. The `wise` column of the output will get them indicating
#'   directions.
#' @return a tibble that has `wise`, `id`, `prop` and the names of the
#'   functions. The `prop` has the proportions of the values satisfied the
#'   condition given by `cond_fun`. Each column with a function name contains 
#'   the output of the function.
#' @importFrom rlang as_function
cond_prop <- function(mat, cond_fun, aggre_funs = NULL, wises = c("Sample", "Protein")) {
  # names are used for column names
  nms <- names(aggre_funs)
  if (!is.null(aggre_funs) && is.null(nms) || any(nms == "")) {
    stop("The names of `aggre_funs` must be given.")
  }
  
  lapply(
    1L:2L,    # row-wise , column-wise
    function(ii) {
      tibble(
        wise = case_when(
          ii == 1L ~ wises[1], 
          ii == 2L ~ wises[2],
          TRUE ~ ""
        ),
        id = dimnames(mat)[[ii]],
        prop = apply(mat, ii, \(x) {
          rlang::as_function(cond_fun)(x) |>  # purrr lambda function
            mean(na.rm = TRUE)
        }),
      ) |> 
        add_column(
          map_dfc(aggre_funs, ~apply(mat, ii, rlang::as_function(.x)))
        )
    }
  ) |> 
    bind_rows() |> 
    mutate(wise = factor(wise, levels = wises))   # fix the order for plots
}

cond_prop_by_prep_type <- function(x, cond_fun, aggre_funs = NULL) {
  to_matrix_by_prep_type(x) |> 
    # compute missing proportion and other statistics
    map(~ cond_prop(.x, cond_fun, aggre_funs)) |>
    bind_rows(.id = "prep_type")   # longitudinal table
}
```

#### Type of missingness

Values can be missing by various reasons in SWATH-MS proteomics data. 
Low abundant proteins are prone to generate missing values due to limit of detection (LOD), though the tendency was found weak in a study ([McGurk _et al._, 2019](https://academic.oup.com/bioinformatics/article/36/7/2217/5650405)).
Such tendency was examined in our data set using the median as a representative of the abundance of a protein or a sample. 

* **Protein-wise investigation revealed inverse correlation** as shown in a scatter plot below. 
Proteins with lower median signals had generally more missing values.
Such trends agree with the assumption of missing at not random.

Please note, because a few samples were not analyzed by both sample preparations, missingness was primarily investigated keeping individual prep. types separate.

```{r Missing proportion, purl=TRUE}
sumstat <- list(
  Median = ~median(.x, na.rm = TRUE) 
  # MAD = ~mad(.x, na.rm = TRUE)
) 

# contains missing proportion together with median, MAD
mssngs <- cond_prop_by_prep_type(spct, is.na, sumstat) |> 
  expss::apply_labels(prop = "The proprotion of missing values")
```

```{r Scatter&Hist - median vs missing proportion}
#| fig.dim = c(10, 8), 
#| fig.cap = "The sample/protein-wise relationship between median intensities and missing value proportions is visualized using a scatter plot in the left panel.
#|   The frequencies of the proportion are presented in histograms together with density plots to the right."

prop_scatter_hist(mssngs, Median, prep_type)
```

Each sample preparation type was separately presented. 

```{r Scatter&Hist - by prep type}
#| fig.dim = c(10, 8), 
for(ii in unique(spct$prep_type)) {
  mssngs |> 
    filter(prep_type == ii) |> 
    prop_scatter_hist(Median, title = ii)
}
```

<br>

####  Sample-wise missingness

##### Remove the samples with almost no data

For the majority of the samples, about a half of the proteins were not measured in individual samples as shown in the histogram below. 
**Two separate clusters** were obvious in the distribution plot of Neat samples' data, 
which implies too much aberration of a few sample(s) with >90% missing values.

```{r Histogram - missing prop of ind. samples, fig.dim = c(6, 3.5)}
for(ii in prep_types) {
  p <- mssngs |> 
    filter(wise == "Sample", prep_type == ii) |> 
    ggplot() +
    geom_histogram(aes(x = prop), binwidth = 0.05, boundary = 1) +
    xlim(0, 1) +
    labs(x = expss::var_lab(mssngs$prop), y = "Frequency") +
    ggtitle(ii)
  print(p)
}
```

```{r * Remove SAMPLES with too many missing values *, purl=TRUE}
too_many_mssng_smpl <- mssngs |> 
  filter(prop > 0.7 & wise == "Sample") |> 
  arrange(desc(prop))
spct <- spct |> 
  anti_join(too_many_mssng_smpl, by = c("sampleid" = "id", "prep_type"))
```

`r english::Words(nrow(too_many_mssng_smpl))` sample(s) had more than 70% missingness.
The ***sample(s) with too many missing values (>70%)*** are listed below and ***removed***. 

```{r Table - samples with too many missing values}
too_many_mssng_smpl |> 
  mutate(prop = sprintf("%1.1f%%", prop * 100)) |>    # to percentage
  select(Sample = id, 
         "Prep. type" = prep_type,
         "Missing proportion" = prop) |> 
  kableExtra::kbl(align = "llrr") |> 
  kableExtra::kable_styling(full_width = F)
```


##### Factors correlated with missingness

Here, potentially correlated other factors were examined.
Because of the significant difference between neat and depleted samples, missing proportion was computed per type as before. 

```{r Missing proportions updated, purl=TRUE}
mssngs_upd <- cond_prop_by_prep_type(spct, is.na, sumstat) |> 
  expss::apply_labels(prop = "The proprotion of missing values")
```

The missing proportion values of those two cohorts were surprisingly different. 
Especially, the missing proportions of IAVI visit 2 depleted samples were prominently higher than others. 

```{r Missing - visit cohort, fig.dim = c(8, 4)}
# Clinical information
load(fn$pr$c01)

# attach clinical information
mssngs_upd_clinc <- mssngs_upd |> 
  filter(wise == "Sample") |> 
  separate(id, c("subjid", "visit_nr"), sep = "-", remove = FALSE) |>
  select(-wise, sampleid = id) |> 
  # add other factors for investigation
  left_join(clinc, by = "subjid") 

ggplot(mssngs_upd_clinc, aes(Cohort, prop)) +
  ggpirate::geom_pirate(aes(color = visit_nr), show.legend = T) +
  labs(color = "Visit number", y = expss::var_lab(mssngs_upd$prop)) +
  facet_wrap(~prep_type)
```

* Missing proportion by Cohort in Depleted samples

```{r, results='markup'}
t.test(prop ~ Cohort, mssngs_upd_clinc, subset = prep_type == "Depleted")
```

* Missing proportion by Cohort in Neat samples

```{r, results='markup'}
tmp <- t.test(prop ~ Cohort, mssngs_upd_clinc, subset = prep_type == "Neat")
print(tmp)
```

The p-value is `r signif(tmp$p.value)`.

* Missing proportion by Cohort in Depleted samples collected at visit 2

```{r, results='markup'}
t.test(prop ~ Cohort, mssngs_upd_clinc, subset = prep_type == "Depleted" & visit_nr == "T2")
```

Focusing on wider spectrum of median values of Neat samples, `Date processed` was found to associate with them. 

```{r Missing - Date processed}
# get Date processed
load(fn$pr$injt)
run_neat <- readxl::read_xlsx(fn$pr$run_neat) |> 
  mutate(FileName = str_replace(FileName, "uL", "ul")) |>   # mistmatch
  left_join(injtn, by = c("FileName" = "run_id")) |>    # sampleid
  mutate(`Date processed` = as.character(`Date processed`)) |> 
  select(sampleid, `Date processed`)

neat <- mssngs_upd_clinc |> 
  left_join(run_neat, by = "sampleid") |> 
  filter(prep_type == "Neat")
  
ggplot(neat, aes(Median, prop)) +
  geom_point(aes(color = `Date processed`)) +
  ylab(expss::var_lab(mssngs_upd$prop)) +
  xlab("Median of `LogIntensities`")
```


```{r Missing - Date processed - box, fig.dim = c(8, 4)}
#| fig.cap = "The number of samples in each column is presented at the bottom"
give.n <- function(x) {
  c(y = min(x), label = length(x))
}

ggplot(neat, aes(`Date processed`, Median)) +
  geom_boxplot(aes(fill = `Date processed`)) +
  geom_jitter(aes(color = `Date processed`)) +
  stat_summary(fun.data = give.n, geom = "text") +
  ylab("Median of `LogIntensities`") +
  theme(axis.text.x = element_blank()) +
  facet_wrap(~Cohort)
```

```{r, fig.dim = c(6, 4), dpi = 96}
mssngs_upd_clinc |> 
  ggplot(aes(Edi, prop)) +
  geom_point(aes(color = prep_type, shape = Cohort)) +
  ylab(expss::var_lab(mssngs_upd$prop)) +
  xlab("Estimated date of infection")
```

#### Protein-wise missingness

##### Remove the proteins with too many missing data (>80%)

```{r * Remove PROTEINS with too many missing values *, purl=TRUE}
too_many_mssng_protein <- mssngs_upd |> 
  filter(prop > 0.8 & wise == "Protein") |> 
  arrange(desc(prop))
spct <- spct |> 
  anti_join(too_many_mssng_protein, by = c(protein_type_id = "id"))
```

`r english::Words(nrow(too_many_mssng_protein))` proteins had more than 80% missingness.
Those ***proteins with too many missing values (>80%)*** are listed below and ***removed***. 

```{r Table proteins with too many missing values}
#  show the proteins with too many missing values
too_many_mssng_protein |> 
  select(protein_type_id = id, prop) |> 
  left_join(proteins, by = "protein_type_id") |>   # get Protein
  mutate(prop = sprintf("%1.1f%%", prop * 100)) |>    # to percentage
  select(Protein, "Prep. type" = prep_type, "Missing proportion" = prop) |> 
  kbl(align = "lrl") |> 
  kable_styling(full_width = F) |> 
  scroll_box(height = "300px")
```

```{r Intermediate status save, echo=FALSE, eval=FALSE, purl=TRUE}
save(spct, file = fn$o$interim)
```

</br>

### Imputation and normalization

```{r FUN - qqplot_per}
#' Normal QQ plot per 
#'
#' @param x input table in longitudinal format
#' @param i_col the columns to use the selection below
#' @param subset selected subset to show
#' @return a ggplot of multiple normal QQ plots
qqplot_per <- function(x, i_col, subset) {
  x |> 
    filter({{i_col}} %in% subset) |> 
    ggplot() +
    geom_qq(aes(sample = LogIntensities), na.rm = T) +
    facet_wrap(enexpr(i_col))
}
```

```{r Imputation preparation}
set.seed(202)
# samples and proteins to show
i_show <- list(
  s = sample(unique(spct$sampleid), 12),
  p = sample(unique(spct$protein_type_id), 12)
)
```

#### Before imputation

The signal intensities were already log-transformed and which of each sample had a normal(-like) distribution. 
The points of obtained data of individual samples lie around a straight line in the normal quantile-quantile plots below.

```{r QQplot of signal before imputation}
qqplot_per(spct, sampleid, i_show$s) +
  ggtitle("Normal Quantile-quantile plot per sample (randomly chosen 12)")
qqplot_per(spct, protein_type_id, i_show$p) +
  ggtitle("Normal QQ plot per protein (randomly chosen 12)")
```


#### Imputation

Each missing value was replaced with a randomly chosen value between one and the minimum of the protein that has the missing value.
This imputation was performed with non-logged values (i.e. 2^`LogIntensities`).

_Here we followed the imputation method suggested by_
_Christofer Karlsson (<christofer.karlsson@med.lu.se>)._

```{r Imputation, purl=TRUE}
### A modified version of Christofer Karlsson's approach
set.seed(20220202)

spct_imp <- lapply(
  split(spct, spct$prep_type),
  function(one_type) {
    # Non-log-transformed intensities
    spct_not_log <- one_type |> 
      pivot_wider(id_cols = sampleid,
                  names_from = protein_type_id,
                  values_from = LogIntensities) |> 
      pivot_longer(-sampleid, names_to = "protein_type_id", values_drop_na = FALSE) |>
      mutate(value = 2^value)
    
    # determine max impute value per protein
    max.impute.value <- spct_not_log |> 
      group_by(protein_type_id) |>
      summarise(max_imp = min(value, na.rm = TRUE), .groups = "drop")
    
    # join the max impute values to the prot.df
    # impute missing values with runif
    prot.df.imputed <- spct_not_log |>
      left_join(max.impute.value, by = "protein_type_id") |>
      mutate(
        Imputed = if_else(is.na(value), TRUE, FALSE),
        r_unif = sapply(max_imp, function(x) {
          runif(1, min = 1, max = x)
        }),
        value = if_else(is.na(value), r_unif, value)
      )
    
    # Store imputed data
    prot.df.imputed |>
      mutate(LogIntensities = log2(value)) |>
      select(sampleid, protein_type_id, LogIntensities)
  }
) |> 
  bind_rows(.id = "prep_type")
```

* Distribution of signal after imputation

```{r QQplot of signal after imputation}
qqplot_per(spct_imp, sampleid, i_show$s) +
  ggtitle("Normal QQ plot per sample (randomly chosen 12)")
qqplot_per(spct_imp, protein_type_id, i_show$p) +
  ggtitle("Normal QQ plot per protein (randomly chosen 12)")
```

#### Normalization

<!-- ```{r spct_qc NormalyzerDE, purl=TRUE, eval=TRUE} -->
<!-- # NormalyzerDE doesn't allow values < 1 -->
<!-- stopifnot(min(spct$LogIntensities) >= 1) -->

<!-- data_mat <- 2^t(long_to_matrix(spct_imp)) -->
<!-- smpl <- distinct(spct_imp, sample = sampleid) |> -->
<!--   separate(sample, c("subjid", "group"), remove = FALSE) |> -->
<!--   select(-subjid) |> -->
<!--   mutate(group = factor(group)) -->
<!-- stopifnot(identical(colnames(data_mat), smpl$sample)) -->

<!-- sum_exp_obj <- SummarizedExperiment::SummarizedExperiment( -->
<!--   data_mat, -->
<!--   colData = smpl, -->
<!--   rowData = data.frame(protein = rownames(data_mat)) -->
<!-- ) -->
<!-- NormalyzerDE::normalyzer(jobName = "HIV", -->
<!--                          experimentObj = sum_exp_obj, -->
<!--                          outputDir = "../cache") -->
<!-- ``` -->


The imputed data was normalized by Cyclic Loess using `normalizeCyclicLoess` function in `limma` package (`r packageVersion("limma")`).
The normalization was performed keeping the data of the two sample types separate.

```{r LOESS normalization, purl=TRUE}
# Normalized by Cyclic Loess
spct_proc <- spct_imp |>
  # per preparation type
  to_matrix_by_prep_type() |>
  map_dfr(
    ~t(.x) |>
      limma::normalizeCyclicLoess() |>
      as_tibble(rownames = "protein_type_id") |>
      pivot_longer(-protein_type_id, names_to = "sampleid", values_to = "LogIntensities")
  )
```

* Distribution of signal after the normalization

```{r QQplot of signal after LOESS}
qqplot_per(spct_proc, sampleid, i_show$s) +
  ggtitle("Normal QQ plot per sample (randomly chosen 12)")
qqplot_per(spct_proc, protein_type_id, i_show$p) +
  ggtitle("Normal QQ plot per protein (randomly chosen 12)")
```


#### Progress of distribution

```{r Plot - progress}
bind_rows(
  spct |> 
    select(sampleid, protein_type_id, LogIntensities) |> 
    filter(sampleid %in% i_show$s[1:3]) |> 
    mutate(Status = "Before"),
  spct_imp |> 
    filter(sampleid %in% i_show$s[1:3]) |> 
    mutate(Status = "After Imputatation"),
  spct_proc |> 
    filter(sampleid %in% i_show$s[1:3]) |> 
    mutate(Status = "After Normalization")
) |> 
  mutate(
    # fix the order
    Status = factor(Status, levels = c(
      "Before", 
      "After Imputatation",
      "After Normalization"
    ))
  ) |> 
  ggplot() +
  geom_histogram(aes(x = LogIntensities), na.rm = T, bins = 30) +
  facet_grid(sampleid ~ Status) +
  xlab("log of signal intensities") +
  ggtitle("Signal distribution change per sample (randomly chosen 3)")
```

```{r Save, echo=FALSE, eval=FALSE, purl=TRUE}
proteins <- proteins %>% 
  filter(protein_type_id %in% spct_proc$protein_type_id)

save(spct_proc, proteins, file= fn$o$sp02)
```

```{r Confirm the same processed output, echo=FALSE}
if (!is.null(fn$check) && file.exists(fn$check)) {
  spct_proc_thru_report <- spct_proc
  load(fn$check, envir = e <- new.env())
  stopifnot(identical(spct_proc_thru_report, e$spct_proc))
  rm(e)
}
```

<br>

## After QC

### Dimension

* Number of samples : `r n_distinct(spct_proc$sampleid)`
* Number of proteins : `r n_distinct(spct_proc$protein_type_id)`

### Proteins

```{r Venndiagram - after QC}
measured_by <- proteins |> 
  filter(protein_type_id %in% spct_proc$protein_type_id) |> 
  mutate(value = TRUE) |> 
  pivot_wider(id_cols = id, names_from = prep_type, values_fill = FALSE)

ggplot(measured_by) +
  ggvenn::geom_venn(aes(A = Neat, B = Depleted)) +
  coord_fixed() +
  theme_void()
```

### Plots

* Mean-Difference plot

```{r MA-plots, fig.dim=c(10, 8)}
#' Make a mean-difference plot
#'
#' @param spct_tbl a data.frame that contains protein data in long format
mean_diff_plot <- function(spct_tbl) {
  id_cols <- quote(c(subjid, protein_type_id))

  # visit number in columns
  by_visit <- spct_tbl |>
    separate(sampleid, c("subjid", "visit_nr"), sep = "-") |> 
    pivot_wider(id_cols = !! id_cols, names_from = visit_nr, values_from = LogIntensities)
  
  # LogIntensities in a matrix
  mat <- select(by_visit, -!! id_cols) |> 
    as.matrix()
  
  # pairwise combintions 
  nm <- combn(colnames(mat), 2, paste, collapse = " & ")
  
  rotation <- do.call(
    "cbind", 
    combn(ncol(mat), 2, FUN = function(x) {
      out <- matrix(0, nrow = ncol(mat), ncol = 2)
      out[x, 1] <- c(1, -1)   # difference
      out[x, 2] <- 1/2        # mean
      out
    }, simplify = F)
  ) |> 
    `colnames<-`(paste(c("Difference", "Mean"), rep(nm, each = 2), sep = "_"))
  
  # Mean, Difference by matrix multiplication
  ma <- add_column(
    select(by_visit, !! id_cols), 
    mat %*% rotation |> 
      as_tibble()
  ) |> 
    pivot_longer(-!! id_cols) |>    # one value per row
    separate(name, c("what", "pair"), sep = "_") |> 
    pivot_wider(names_from = what) |>     # split to mean and difference
    drop_na(Difference, Mean) |> 
    left_join(select(clinc, subjid, Cohort), by = "subjid") |>   # add Cohort
    separate(protein_type_id, c("protein_id", "prep_type"), sep = "-") # prep_type
  for(ii in unique(ma$pair)) {
    p <- ma |> 
      filter(pair == ii) |> 
      ggplot(aes(Mean, Difference)) +
      geom_point() +
      geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
      facet_wrap(Cohort ~ prep_type) +
      ggtitle(paste("Visits", ii))
    print(p)
  }
}

mean_diff_plot(spct_proc)
```


