```{r VLvar}
var <- "vload"
varid <- "VL"
varname="Viral Load"
```

```{r VLspline}
lambda <- AHIprot::boxcoxlambda(CV[[varid]], "day", var, "subjid")
#lambda$y <- 1 
CVsplines[[varid]] <- splines(CV[[varid]], "day", var, "subjid", lambda)
tr <- function(l) {
  if (l==0)
    "log10-transformed"
  else if (l==1)
    "untransformed"
  else
    paste0("x^", l)
}
```

```{r VLdaydistr, fig.cap=sprintf("Distribution of %s and time points before and after transformation, colored according to subjid. %s is %s and day is %s (log10(day+1)).", varname, varname, tr(CVsplines[[varid]]$lambda$y), tr(CVsplines[[varid]]$lambda$x)), warning=FALSE}
CVsplines[[varid]]$pl$distr
```

An optimal smoothing parameter (`spar`) is calculated using leave-one-out (LOO) cross-validation (CV). A smoothing spline is calculated per patient, but the CV sum of squares (SS) is computed over all patients. The value of `spar` in the range $[0,1]$ that give the minimum SS is selected.

```{r VLspar, fig.cap="Sum of squares versus spar for Viral Load.", out.width="50%"}
CVsplines[[varid]]$pl$spar + theme_bw()
```

```{r VLbeforeART0, fig.width=10, fig.height=10, fig.cap="VL before start of ART. The smoothing spline is shown in red.", warning=FALSE}
VL0 %>% ggplot(aes(x=day, y=vload)) + geom_point() + geom_line(data=CVsplines[[varid]]$spline, color="red") + scale_y_log10() + facet_wrap(~subjid) + theme_bw() + xlim(c(0,1096))
```

```{r VLbeforeART, fig.width=10, fig.height=10, fig.cap="Viral Load before start of ART. The smoothing spline is shown in red.", warning=FALSE}
CV[[varid]] %>% ggplot(aes(x=day, y=!!as.symbol(var))) + geom_point() + geom_line(data=CVsplines[[varid]]$spline, color="red") + scale_x_continuous(limits=c(NA,1096), trans = scales::boxcox_trans(CVsplines[[varid]]$lambda$x, 1), breaks=c(30, 61, 183, 365, 730, 1096), labels = c("1m", "2m", "6m", "1y", "2y", "3y")) + scale_y_continuous(trans = scales::boxcox_trans(CVsplines[[varid]]$lambda$y), breaks=10^seq(2,8,2), labels = c("100", "1e4", "1e6", "1e8")) + facet_wrap(~subjid) + theme_bw()  
```

```{r NVL, fig.cap="Number of patients with Viral Load observation per time point.", fig.width=14}
data.frame(day=CVsplines[[varid]]$splw$day, n=rowSums(!is.na(CVsplines[[varid]]$splw[, -1]))) %>% ggplot(aes(x=day, y=n)) + geom_point() + scale_x_continuous(breaks=c(42, 100, round(seq(365.25, 3000, 365.23))), trans=scales::boxcox_trans(0, 1))
```

\FloatBarrier

### Correlation between profiles

#### All time points

Calculate pairwise correlation between patients VL curves, will be based on different number of data points depending on how long VL was observed for that particular patient.

```{r VLsplinecluster, fig.width=10, fig.height=10, fig.cap="Hierarchical clustering based on Pearson correlation of log10(VL) before start of ART. Clusters of VL profiles..", fig.show="hold", fig.width=10, fig.height=10, out.width="50%"}
rho <- cor(CVsplines[[varid]]$splw[,-1], use="pair") 
h <- hclust(as.dist(1-rho))
plot(h)
cl <- cutree(h, k=4)
cl <- data.frame(subjid=names(cl), cl=cl)
 
CVsplines[[varid]]$spline %>% left_join(cl) %>% ggplot(aes(x=day, y=vload, group=subjid, color=paste("Cluster ", cl))) + geom_line() + scale_y_log10() + facet_wrap(~paste(cl, subjid, sep=": ")) + theme_bw() + scale_color_discrete("")
```

#### Within 36 months (1096 days)

```{r VLclusters1096, fig.cap="Correlation based clusters of VL profiles, all samples before day 1096 and before ART.", fig.width=10, fig.height=10, fig.show="hold", out.width="50%"}
rho <- cor(CVsplines[[varid]]$splw %>% filter(day<1096) %>% dplyr::select(-day), use="pair")
h <- hclust(as.dist(1-rho))
plot(h)
cl <- cutree(h, k=4)
cl <- data.frame(subjid=names(cl), cl=cl)
 
CVsplines[[varid]]$spline %>% filter(day<1096) %>% left_join(cl) %>% ggplot(aes(x=day, y=vload, group=subjid, color=paste("Cluster ", cl))) + geom_line() + scale_y_log10() + facet_wrap(~paste(cl, subjid, sep=": ")) + theme_bw() + scale_color_discrete("")
```

### Cluster profiles based on Euclidean after center

```{r VLsplinekomb, fig.width=10, fig.height=10, fig.cap="Hierarchical clustering based on mean and euclidean distance of centered log10(VL) before start of ART. Clusters of VL profiles..", fig.show="hold", fig.width=10, fig.height=10, out.width="50%"}
lmt <- c(30,364)
z <- CVsplines[[varid]]$splw %>% dplyr::filter(day>=lmt[1], day<=lmt[2]) %>% dplyr::select(which(colMeans(is.na(.))==0))

m <- colMeans(z[, -1])

zcenter <- scale(z[, -1], center=TRUE, scale=FALSE)
d <- dist(t(zcenter))#/sqrt(nrow(z))
dm <- dist(m)

h <- hclust(d+dm)
plot(h)
cl <- cutree(h, 2)
cl <- data.frame(subjid=names(cl), cl=cl)
CVsplines[[varid]]$spline %>% left_join(cl) %>% filter(!is.na(cl), day>=lmt[1], day<=lmt[2]) %>% ggplot(aes(x=day, y=vload, group=subjid, color=paste("Cluster ", cl))) + geom_vline(xintercept = 30) + geom_vline(xintercept=364) + geom_line() + theme_bw() + scale_color_discrete("") + scale_x_continuous(trans=scales::boxcox_trans(CVsplines[[varid]]$lambda$x, 1)) + scale_y_continuous(trans=scales::boxcox_trans(CVsplines[[varid]]$lambda$y)) 
```


\FloatBarrier



### Cluster Viral Load profiles based on Euclidean distance


Based on the cubic spline predicted Viral Load at evenly spread (on transformed scale) time points the Euclidean distance between patients Viral Load curves can be calculated. As no extrapolation is done, no value will be availbale for time points outside the observed range. Hence, the number of patients that can be compared depend on the studied time interval.

For a given time interval, all patients with observations at the beginning and end of this interval are clusterd based on their Euclidean distance using complete linkage hierarchical clustering. The optimal number of clusters is determined using the Silhouette value.

```{r EuclideanclustersVL, warning=FALSE}
timewindows <- as.list(as.data.frame(t(cbind(min=c(rep(365.25/12*1:6, 2), rep(365.25/12*1:3, 5:3)), max=c(rep(365.25*3,6), rep(365.25,6), 365.25/12*c(2:6,3:6, 4:6))))))
names(timewindows) <- sapply(timewindows, function(lmt) paste0(paste(lmt/365.25*12, collapse="-"), "m"))
euclust[[varid]] <- sapply(names(timewindows), function(lmtnm) {
  Eucluster(CVsplines[[varid]]$splw, CVsplines[[varid]]$spline, CVsplines[[varid]]$lambda, timewindows[[lmtnm]], lmtnm, time="day", value=var)
}, USE.NAMES = TRUE, simplify = FALSE)
```


```{r}
##Observed days
mnmx <- CV[[varid]] %>% group_by(subjid) %>% summarise(mn=min(day), mx=max(day))
```

```{r VLclusters, out.height=sprintf('%i%%', round(100/length(timewindows))), fig.show='hold', fig.cap=sprintf('%s clusters. Complete linkage hierarchical clustering based on Euclidean distance. Only patients with %s data during the studied time period are included in the clustering.', varname, varname), eval=FALSE}
for (lmt in names(timewindows)) {
  plot(euclust[[varid]][[lmt]]$hpl) 
  plot(euclust[[varid]][[lmt]]$pl)
}
```

```{r VLclusters0, out.width='50%', fig.show='hold', fig.cap='VL, complete linkage hierarchical clustering based on Euclidean distance. Only patients with VL data during studied time period are included in the clustering.'}
plot(euclust$VL[["1-36m"]]$hpl)
plot(euclust$VL[["1-36m"]]$pl)
plot(euclust$VL[["6-36m"]]$hpl)
plot(euclust$VL[["6-36m"]]$pl)
plot(euclust$VL[["1-6m"]]$hpl)
plot(euclust$VL[["1-6m"]]$pl)
plot(euclust$VL[["1-12m"]]$hpl)
plot(euclust$VL[["1-12m"]]$pl)
```


\FloatBarrier